---
title: "Section 4: Population Scale Analysis"
author: Nicole Chang
date: 2023-05-16
format: gfm
editor_options: 
  chunk_output_type: inline
---

## Q13. Read this file into R and determine the sample size for each genotype and their

corresponding median expression levels for each of these genotypes.

Reading file:

```{r}
experiment <- read.table("https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt")
View(experiment)
```

Determine sample size:

```{r}
nrow(experiment)
```

Looking at each genotype:

```{r}
table(experiment$geno)
```

Separating each genotype:

```{r}
AA <- experiment[experiment[,"geno"]=="A/A",]
AG <- experiment[experiment[,"geno"]=="A/G",]
GG <- experiment[experiment[,"geno"]=="G/G",]
```

Median expression level for A/A genotype:

```{r}
median(AA$exp)
```

Median expression level for A/G genotype:

```{r}
median(AG$exp)
```

Median expression level for G/G genotype:

```{r}
median(GG$exp)
```

## Q14. Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Load ggplot2

```{r}
library(ggplot2)
```

Create boxplot

```{r}
ggplot(experiment) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```

Based on the plot, the relative expression of G/G shows a significant reduction compared to the relative expression of A/A in this gene. The SNP seems to affect the expression of ORMDL3.
